---
title: "AR61b CTD profiles"
format: html
---

Analyse and plot AR61b CTD data.
CTD cast data are downloaded from the NES-LTER API
and joined to bottle chlorophyll measurements. This could be done in R using 
functions from dplyr and tidyr, but I've done this here in python:

https://github.com/blongworth/whoi-ctd-data

# Setup libraries

We'll use functions from dplyr, tidyr, and ggplot. It's easy to load these
and other packages at once using `library(tidyverse)`. Is this a good idea?
It's easy, but makes it harder to know where functions come from.

We'll also use the `oce` package later in the notebook, so we'll load it here.

`oce` can be installed with `install.packages("oce")`.

It's good practice to load all libraries
at the beginning of the code.

```{r}
library(tidyverse)
library(oce)
```

Note that `library(tidyverse)` tells us what packages we loaded,
and more importantly, what name conflicts we've created.

Think of package functions like any other variable or object (they are!).

If we do `x <- 4` and then `x <- 2`, `x` is now 2
because reassigning overwrites the previous value of `x`.

# Load data

We'll load the data directly from a url on github so we don't have to download it.
`read_csv()` from the tidyverse `readr` package can do this directly.

If we downloaded the data manually, 
we'd put it within a `data` folder within our project.

```{r}
df <- read_csv("https://raw.githubusercontent.com/blongworth/whoi-ctd-data/main/data/ar61b_ctd_cast_data.csv")
```

# Explore!

Check out the structure (also use data browser from environment or `view()`)

```{r}
str(df)
```

First few lines...

```{r}
head(df)
```

Are the data types correct? Is this important for our analysis?

* continuous variables like temp, sal, ox, and chl are good
as floating point (double)
* cast and niskin could be integers
* cast, niskin, station could be factors
* station will be in alphabetical order
* `niskin bottle` could be renamed as `niskin_bottle` or `niskin`
to make it easier to use without backticks.

Is any of this necessary? No. 
Could it make the analysis easier and less error prone? Maybe.

## Investigate stations

How many stations are there? What are they called?

```{r}
unique(df$station)
length(unique(df$station))
```

Which station had the most casts?

```{r}
# base R
table(df$station)

# dplyr
df |> count(station) |> arrange(desc(n))
```

This is number of rows in the data for each station.
How do we get number of casts?

`distinct()` will take only the first row for each station/cast combo.

```{r}
df |>
  distinct(station, cast) |>   # ensure one row per station–cast combo
  count(station, name = "n_casts") |> 
  arrange(desc(n_casts))
```

# Station map

Let's see where these stations are! This just plots lat and lon of points.
If we want a better map, there are a bunch of ggplot and other methods
for producing nice maps with land, grids, bathymetry, etc.

```{r}
df |> 
  ggplot(aes(longitude, latitude, color = station)) +
  geom_point()
```

Can we fix the station order?

```{r}
lvl_order <- c("L1", "L2", "L3", "L4", "L5", "L6", "L7", "L7.5",
               "L8", "L9", "u9a", "u9b", "d9a", "L9.5", "L10", 
               "L11", "u11a", "L12", "L13")

# make an ordered factor
df$station <- factor(df$station, levels = lvl_order, ordered = TRUE)
```

Let's at least make the lat and lon axes have the same interval...

```{r}
df |> 
  ggplot(aes(longitude, latitude, color = station)) +
  geom_point() +
  coord_fixed()
```

Ugly, but makes more sense!

A slightly fancier map...

```{r}
library(maps)   # for world map data

world <- map_data("world")

df |>
  ggplot(aes(x = longitude, y = latitude, color = station)) +
  geom_polygon(
    data = world,
    aes(x = long, y = lat, group = group),
    inherit.aes = FALSE,
    fill = "grey80",
    color = "grey60",
    linewidth = 0.2
  ) +
  geom_point(size = 2) +
  coord_quickmap(xlim = c(-72, -69.5), ylim = c(39, 42)) +
  theme_minimal() +
  labs(x = NULL, y = NULL, color = "Station")

```


# properties vs depth

Let's start checking out CTD profiles!

## temp vs depth

```{r}
df |> 
  ggplot(aes(depth, temperature)) +
  geom_point()
```

Flip axes around to make this look oceanographic!

```{r}
ggplot(data = df, mapping = aes(x = depth, y = temperature)) +
  geom_point(alpha = 0.3) +
  scale_x_reverse() +
  coord_flip() 
```

What's up with that difference in temp in the mixed layer?

Let's zoom in.
To do this, we will use pipes and `dplyr::filter()` 
to filter by depth before piping the filtered data to ggplot.

```{r}
df |> 
  filter(depth < 100) |> 
  ggplot(aes(depth, temperature)) +
  geom_point(alpha = 0.3) +
  scale_x_reverse() +
  coord_flip()
```

We can also make a univariate plot to look at the difference in population.

Maybe a histogram?

```{r}
df |> 
  filter(depth < 100) |> 
  ggplot(aes(temperature)) +
  geom_histogram()
```


What's different about these two populations? Is it separated by station?


```{r}
df |> 
  filter(depth < 100) |> 
  ggplot(aes(depth, temperature, color = station)) +
  geom_point(alpha = 0.3) +
  scale_x_reverse() +
  coord_flip()
```

Looks like it's grouped by station. Maybe it's separated by latitude. A front?

```{r}
df |> 
  filter(depth < 100) |> 
  ggplot(aes(depth, temperature, color = latitude)) +
  geom_point(alpha = 0.3) +
  scale_x_reverse() +
  coord_flip()
```

Still hard to see. What about making latitude groups? 

We'll make a column splitting data into high and low latitudes
using `dplyr::mutate()` and ifelse to conditionally assign 
high and low based on the value of the latitude column

```{r}
df |> 
  filter(depth < 100) |> 
  mutate(lat_grp = ifelse(latitude > 40.4, "high", "low")) |> 
  ggplot(aes(depth, temperature, color = lat_grp)) +
  geom_point(alpha = 0.3) +
  scale_x_reverse() +
  coord_flip()
```

Getting there, but there's still some southern stuff with high mixed layer temp.
Let's try mapping the temp to see what's going on.

```{r}
df |> 
  filter(depth < 100) |> 
  ggplot(aes(longitude, latitude, color = temperature)) +
  geom_point()
```

Let's change to a better temperature scale

```{r}
df |> 
  filter(depth < 100) |> 
  ggplot(aes(longitude, latitude, color = temperature)) +
  geom_point() +
  scale_color_viridis_c(option = "inferno")
```

Adjust colors to make things clear.
We're using mutate and ifelse again to make temp into a TRUE/FALSE

```{r}
df |> 
  filter(depth < 100) |> 
  mutate(temp_grp = ifelse(temperature > 17, "high", "low")) |> 
  ggplot(aes(longitude, latitude, color = temp_grp)) +
  geom_point()
```

Looks like the front is around 40.1N. Let's add this to our plot.
We can iterate on the exact latitude of the front until 
we think we've found the right value.

```{r}
df |> 
  filter(depth < 100) |> 
  mutate(lat_grp = ifelse(latitude > 40.14, "high", "low")) |> 
  ggplot(aes(depth, temperature, color = lat_grp)) +
  geom_point(alpha = 0.3) +
  scale_x_reverse() +
  coord_flip()
```

Now that we have the groups well separated, 
we could compare the groups using a boxplot.

```{r}
df |> 
  filter(depth < 100) |> 
  mutate(lat_grp = ifelse(latitude > 40.14, "high", "low")) |> 
  ggplot(aes(lat_grp, temperature)) +
  geom_boxplot() 
```

## individual plots

Do other measurements also show the front?

What about oxygen and chlorophyll?

```{r}
df |> 
  filter(depth < 100) |> 
  mutate(lat_grp = ifelse(latitude > 40.14, "high", "low")) |> 
  ggplot(aes(depth, chlorophyll, color = lat_grp)) +
  geom_point(alpha = 0.3) +
  scale_x_reverse() +
  coord_flip()
```


```{r}
df |> 
  filter(depth < 100) |> 
  mutate(lat_grp = ifelse(latitude > 40.14, "high", "low")) |> 
  ggplot(aes(depth, oxygen, color = lat_grp)) +
  geom_line(aes(group = cast)) +
  geom_point(alpha = 0.3) +
  scale_x_reverse() +
  coord_flip()
```

### Comparing properties across the front using `pivot_longer()`

Using `pivot_longer()` to compare multiple properties at once and eliminate code repetition:

```{r}
# Reshape to long format for easier comparison across properties
df |> 
  filter(depth < 100) |> 
  mutate(lat_grp = ifelse(latitude > 40.14, "north (cold)", "south (warm)")) |>
  pivot_longer(
    cols = c(temperature, chlorophyll, oxygen),
    names_to = "property",
    values_to = "value"
  ) |>
  ggplot(aes(depth, value, color = lat_grp)) +
  geom_point(alpha = 0.3) +
  scale_x_reverse() +
  coord_flip() +
  facet_wrap(~property, scales = "free_x") +
  labs(title = "Water properties across the front",
       color = "Water mass")
```

Is chlorophyll related to temp or the front?

```{r}
df |> 
  filter(depth < 100) |> 
  mutate(lat_grp = ifelse(latitude > 40.14, "high", "low")) |> 
  ggplot(aes(temperature, chlorophyll, color = lat_grp)) +
  geom_point()
```

Two observations: chlorophyll is pretty low and constant in the warmer,
offshore water, and it looks like chlorophyll is inversely related to temperature
in the colder, more northerly water.

Can we plot a fit line for the onshore water?

```{r}
df |> 
  filter(depth < 100) |> 
  mutate(lat_grp = ifelse(latitude > 40.14, "high", "low")) |> 
  ggplot(aes(temperature, chlorophyll, color = lat_grp)) +
  geom_smooth() +
  geom_point()
```

## Summary statistics across the front using `pivot_longer()` and `pivot_wider()`

Create a summary table comparing mean and standard deviation of properties by water mass and depth:

```{r}
df |> 
  filter(depth < 100) |> 
  mutate(lat_grp = ifelse(latitude > 40.14, "north (cold)", "south (warm)"),
         depth_bin = cut(depth, breaks = seq(0, 100, 50))) |>
  pivot_longer(
    cols = c(temperature, chlorophyll, oxygen, salinity),
    names_to = "property",
    values_to = "value"
  ) |>
  group_by(lat_grp, depth_bin, property) |>
  summarise(mean = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE),
            .groups = "drop") |>
  pivot_wider(
    names_from = property,
    values_from = c(mean, sd),
    names_sep = "_"
  )
```

Let's make a TS plot! There's a great library for oceanographic stuff:
`oce`. We can install it using `install.packages("oce").

oce is pretty picky about data formatting, so we'll have to make our columns
into an oce ctd object using `as.ctd()`.
We're also using `oce::swPressure() to convert depth to pressure.

```{r}
library(oce)

ctd <- as.ctd(salinity = df$salinity,
              temperature = df$temperature,
              pressure = swPressure(df$depth, latitude = 40))

plotTS(ctd,
       inSitu = FALSE,      # potential temp (default)
       nlevels = 8,         # number of isopycnals
       col.rho = "gray40",  # color of density lines
       drawFreezing = TRUE) # show freezing line

```

This is everything. Do the TS plots look different for the warm and cold masses?

```{r}
df_north <- df |> 
  filter(latitude > 40.14)

df_south <- df |> 
  filter(latitude <= 40.14)
```

North (cold) water

```{r}
ctd_north <- as.ctd(salinity = df_north$salinity,
              temperature = df_north$temperature,
              pressure = swPressure(df_north$depth, latitude = 40))

plotTS(ctd_north,
       inSitu = FALSE,      # potential temp (default)
       nlevels = 8,         # number of isopycnals
       col.rho = "gray40",  # color of density lines
       drawFreezing = TRUE) # show freezing line

```

South (warmer) water

```{r}
ctd_south <- as.ctd(salinity = df_south$salinity,
              temperature = df_south$temperature,
              pressure = swPressure(df_south$depth, latitude = 40))

plotTS(ctd_south,
       inSitu = FALSE,      # potential temp (default)
       nlevels = 8,         # number of isopycnals
       col.rho = "gray40",  # color of density lines
       drawFreezing = TRUE) # show freezing line

```

Does density structure across the front show differences?
We'll calcuate sigma theta and add it to the original df.
Should we add to the original df or to a new dataframe? 
How might reassigning df cause issues?

```{r}
# Calculate sigma-theta and add to dataframe
df <- df |>
  mutate(
    sigma_theta = swSigmaTheta(salinity, temperature, pressure = swPressure(depth, latitude = latitude))
  )
```

Let's plot it and add some nice labels.

```{r}
# Plot sigma-theta vs depth
df |> 
  ggplot(aes(sigma_theta, depth)) +
  geom_point(alpha = 0.3) +
  scale_y_reverse() +
  labs(x = "σ-θ (kg/m³)", y = "Depth (m)", 
       title = "Potential Density Anomaly vs Depth")
```

Is stratification different at different stations?

```{r}
# Color by station to see stratification differences
df |> 
  ggplot(aes(sigma_theta, depth, color = station)) +
  geom_point(alpha = 0.3) +
  scale_y_reverse() +
  facet_wrap(~station) +
  labs(x = "σ-θ (kg/m³)", y = "Depth (m)")
```

Is stratification different across the front?

```{r}
# Compare north/south front with density
df |> 
  mutate(lat_grp = ifelse(latitude > 40.14, "north (cold)", "south (warm)")) |>
  mutate(
    sigma_theta = swSigmaTheta(salinity, temperature, pressure = swPressure(depth, latitude = latitude))
  ) |>
  ggplot(aes(sigma_theta, depth, color = lat_grp)) +
  geom_point(alpha = 0.3) +
  scale_y_reverse() +
  labs(x = "σ-θ (kg/m³)", y = "Depth (m)", color = "Water mass",
       title = "Density structure across the front")
```

# For more info

Check out the NES-LTER API!

https://nes-lter-api.whoi.edu/

There's also a great R notebook demonstrating getting data directly from the API
and making more cool plots:

https://colab.research.google.com/drive/1F9m-WSOyB3LA9IqvWDKwhQVPrBG_c5Sz?usp=sharing
