---
title: "AR66 CTD profiles"
format: html
---

Analyse and plot AR66 CTD data.
CTD cast data are downloaded from the NES-LTER API
and joined to bottle chlorophyl measurements

# Setup libraries

```{r}
library(tidyverse)
```

# Load data

```{r}
df <- read_csv("data/ar61b_ctd_cast_data.csv")
```

# Explore!

Check out the structure (also use data browser from environment or `view()`)

```{r}
str(df)
```

First few lines

```{r}
head(df)
```

Are the data types correct? Is this important for our analysis?

How many stations are there? What are they called?

```{r}
unique(df$station)
length(unique(df$station))
```

Which station had the most casts?

```{r}
# base R
table(df$station)

# dplyr
df |> count(station) |> arrange(desc(n))
```
This is number of rows. How do we get number of casts?

```{r}
df |>
  distinct(station, cast) |>   # ensure one row per station–cast combo
  count(station, name = "n_casts") |> 
  arrange(desc(n_casts))
```

# Station map

```{r}
df |> 
  ggplot(aes(longitude, latitude, color = station)) +
  geom_point()
```

# properties vs depth

plot temp vs depth

```{r}
df |> 
  ggplot(aes(depth, temperature)) +
  geom_point()
```

```{r lm-fit, fig.alt="Scatter plot of life expectancy vs GDP per capita with a blue trend line summarising the relationship between variables, and gray shaded area indicating 95% confidence intervals for that trend line."}
ggplot(data = df, mapping = aes(x = depth, y = temperature)) +
  geom_point(alpha = 0.3) +
  scale_x_reverse() +
  coord_flip() +
  geom_smooth()
```

What's up with that difference in temp in the mixed layer?

Let's zoom in.

```{r}
df |> 
  filter(depth < 100) |> 
  ggplot(aes(depth, temperature)) +
  geom_point(alpha = 0.3) +
  scale_x_reverse() +
  coord_flip()
```

What's different about these two populations?


```{r}
df |> 
  filter(depth < 100) |> 
  ggplot(aes(depth, temperature, color = station)) +
  geom_point(alpha = 0.3) +
  scale_x_reverse() +
  coord_flip()
```
Looks like it's grouped by station. Maybe it's separated by latitude. A front?


```{r}
df |> 
  filter(depth < 100) |> 
  ggplot(aes(depth, temperature, color = latitude)) +
  geom_point(alpha = 0.3) +
  scale_x_reverse() +
  coord_flip()
```

Still hard to see. What about making latitude groups?

```{r}
df |> 
  filter(depth < 100) |> 
  mutate(lat_grp = ifelse(latitude > 40.4, "high", "low")) |> 
  ggplot(aes(depth, temperature, color = lat_grp)) +
  geom_point(alpha = 0.3) +
  scale_x_reverse() +
  coord_flip()
```

Getting there, but there's still some southern stuff with high mixed layer temp.
Let's try mapping the temp to see what's going on.


```{r}
df |> 
  filter(depth < 100) |> 
  ggplot(aes(longitude, latitude, color = temperature)) +
  geom_point()
```

Let's adjust colors to make things clear.


```{r}
df |> 
  filter(depth < 100) |> 
  mutate(temp_grp = ifelse(temperature > 17, "high", "low")) |> 
  ggplot(aes(longitude, latitude, color = temp_grp)) +
  geom_point()
```

Looks like the front is around 40.1N. Let's add this to our plot.

```{r}
df |> 
  filter(depth < 100) |> 
  mutate(lat_grp = ifelse(latitude > 40.14, "high", "low")) |> 
  ggplot(aes(depth, temperature, color = lat_grp)) +
  geom_point(alpha = 0.3) +
  scale_x_reverse() +
  coord_flip()
```

Do other measurements also show the front?

What about oxygen and chlorophyl?


```{r}
df |> 
  filter(depth < 100) |> 
  mutate(lat_grp = ifelse(latitude > 40.14, "high", "low")) |> 
  ggplot(aes(depth, chlorophyl, color = lat_grp)) +
  geom_point(alpha = 0.3) +
  scale_x_reverse() +
  coord_flip()
```


```{r}
df |> 
  filter(depth < 100) |> 
  mutate(lat_grp = ifelse(latitude > 40.14, "high", "low")) |> 
  ggplot(aes(depth, oxygen, color = lat_grp)) +
  geom_point(alpha = 0.3) +
  scale_x_reverse() +
  coord_flip()
```
